---
layout: post
title: "Increasing (non-) science and maths participation"
comments: true
tags: scientific-journalism education mathematics
---

This is a really important topic: mathematics and science participation has been
decreasing for a while, and an article in
[The Conversation](http://bit.ly/1LhqMJg) (by Rachel Wilson) provides some data
as evidence.

I have a couple of gripes however, with the data as it is presented. First,
let's look at the figure (reproduced from the author's data) for mathematics:

```{r,width=6,height=6,echo=FALSE,res=96,fig.path='{{site.baseurl}}public/img/back-img/'}
library(ggplot2)
dat <- read.csv("../NonPostFiles/data-giV57.csv", header = TRUE)
dat <- dat %>% mutate(Gender = factor(Gender, levels = unique(Gender)))
pl1 <- ggplot(dat %>% filter(Year %in% c(2001, 2014), Variable == "No maths"),
aes(x = interaction(Gender, Year), y = Value, fill = Gender))
print(pl1 + geom_bar(stat = "identity") + xlab("Gender/Year") + ylim(c(0, 50)) +
ylab("Proportion"))
```

Firstly, the title of the figure in The Conversation's piece, was **Declining
science and maths participation in schools**. The figure was likely made by the
editorial team, however the figure shows an *increase* in values. This is
because the figure actually plots the percentage of ATAR eligible students
**not** studying mathematics. Clearly the figure needs better a better title, or
the numbers need to be reversed, e.g.:

```{r,width=6,height=6,echo=FALSE,fig.path='{{site.baseurl}}public/img/back-img/'}
pl2 <- ggplot(dat %>% filter(Year %in% c(2001, 2014), Variable == "No maths"),
aes(x = interaction(Gender, Year), y = 100 - Value, fill = Gender))
print(pl2 + geom_bar(stat = "identity") + xlab("Gender/Year") +
ylab("Proportion"))
```

I think that this is a much better representation of the thrust of the argument
in the article, that participation in mathematics for ATAR eligible students has
declined.

However, whilst the article is very clear about the decline between the years
2001 and 2014, there is no discussion of the fact that it appears this decline
arrested in around 2009. Looking at the full dataset (which the author links to
in the [published version](http://www.maths.usyd.edu.au/u/SMS/MMW2015.pdf),
co-authored by John Mack) shows this in more detail:

```{r,width=6,height=6,echo=FALSE,fig.path='{{site.baseurl}}public/img/back-img/'}
pl3 <- ggplot(dat %>% filter(Variable == "No maths"), aes(x = Year, y = 100 - Value,
colour = Gender))
print(pl3 + geom_line() + geom_point() + xlab("Year") + ylab("Proportion"))
```

There even appears to be an *increase* in participation amongst males since
around 2009! Given this report is an update from one that included 2001 to 2011
data, this levelling off of the decline would have been apparent in the earlier
report ([found here](http://www.maths.usyd.edu.au/u/SMS/MWW2013.pdf)), and I
think should have reported with more earnest.

So what do we get out of this? The decline in participation is serious, but the
levelling off is great if it continues (or participation increases), titles of
figures should match the data they present, and always present the whole data
(not just a cherry-picked version of it).
