---
layout: post
title: "Majority rules: counting a winner"
comments: true
tags: voting R
---

Most of us have had to vote at some stage: in primary/high school, perhaps to
elect class/school captains, during university to choose the best presentation
in your physics class, and obviously if you're old enough, voting in elections
for state/country leaders.

In Australia, we use
[preferential voting](http://en.wikipedia.org/wiki/Ranked_voting_system) to
elect officials, and for the lower house (House of Representatives), counting is
done using a full preferential system, also known as instant runoff voting
(IRV).

That's all well and good, and you can read up more about preferential voting at
the linked Wikipedia page, but why am I discussing voting? I was recently
involved in a judging session where nine judges (myself included) ranked five
possible candidates, in preferential order.

When it came time to count the votes, there was not a clear winner (actually,
there was after looking at the data afterwards, but it didn't make a
difference!), so the question became how to decide on the winner? The chosen
method: sum the preferences for each candidate, with the candidate that had the
*smallest* number of total preferences being the winner.

As a statistician, this irked me. We had data in the form of ranks, but were
treating them as continuous data. Just because two of the judges placed
Candidate A above Candidate B, doesn't mean the difference between the two
candidates was judged the same, rather that both judges *preferred* Candidate A
to Candidate B.

Summing, or taking the mean of the preferences, is known as the
[Borda count](http://en.wikipedia.org/wiki/Borda_count), and by the description
of how it's done (mean/sum), you can see that it is an *average* method: the
candidate that wins the vote is often the consensus view, not that which is
preferred by the majority.

So what other methods could we use? The preferential method as described above?
Or the [Condorcet method](http://en.wikipedia.org/wiki/Condorcet_method), which
is a method that chooses the candidate that would win by majority against
pairings with all the other candidates. The Condorcet method is a *majoritan*
method. Unfortunately it doesn't always work, because you can get cycles. A
great analogy is on the previously linked Wikipedia page: there may be a
situation where each candidate wins against another, like in a
rock-paper-scissors game. So you need some way of breaking ties.

So how do the methods compare for the session I was judging? Let's have a look!
Firstly, I found some code to calculate both the Borda and Condorcet winner
[here](http://www.r-bloggers.com/condorcet-ranking-and-rcpp/). And here's the
data:

```{r,eval=TRUE}
votes <- structure(list(Candidate = c("A", "B", "C", "D", "E"),
                        A = c(1L, 3L, 2L, 4L, 5L), B = c(3L, 4L, 2L, 1L, 5L),
                        C = c(2L, 4L, 5L, 1L, 3L), D = c(2L, 4L, 1L, 3L, 5L),
                        E = c(3L, 4L, 5L, 2L, 1L), F = c(1L, 3L, 2L, 4L, 5L),
                        G = c(1L, 3L, 5L, 4L, 2L), H = c(3L, 2L, 5L, 4L, 1L),
                        I = c(4L, 1L, 5L, 2L, 3L)),
                   .Names = c("Candidate", paste("J", 1:9, sep = "")),
				   class = "data.frame", row.names = c(NA, -5L))
```

Firstly, what's the Borda count winner?

```{r,echo=FALSE,eval=TRUE}
AvgRank <- function(BallotMatrix) {
    Ballots <- as.matrix(BallotMatrix[, -1], mode = "numeric")
    Num_Candidates <- dim(Ballots)[1]
    Names <- BallotMatrix[, 1]
    Ballots[is.na(Ballots)] <- Num_Candidates + 1  #Treat blanks as one worse than min
    MeanRanks <- rowMeans(Ballots)
    Rankings <- data.frame(Names, MeanRanks)
    Rankings <- Rankings[order(rank(Rankings[, 2], ties.method = "random")),
        ]  #Ties handled through random draw
    Rankings <- data.frame(Rankings, seq_along(Rankings[, 1]))
    names(Rankings) <- c("Names", "Average Rank", "Position")
    return(Rankings)
}
AvgRank(votes)
```

Candidate A is the winner by the Borda count, follwed by Candidate D. What if we
do the full preferential count? There were 9 judges, so after distributing
preferences, a candidate needs 5 votes to be declared a winner. I did this
manually, and came up with again, Candidate A, followed by Candidate D.

Finally, how about the Condorcet method? Recall this method counts pairwise
winners:
```{r,echo=FALSE,eval=TRUE}
VoteExtract <- function(BallotMatrix) {
    Votes <- as.matrix(BallotMatrix[, -1], mode = "numeric")
    Num_Candidates <- dim(Votes)[1]
    Votes[is.na(Votes)] <- Num_Candidates + 1  #Treat blanks as one worse than min
    return(Votes)
}

PairCount <- function(Votes) {
    Num_Candidates <- dim(Votes)[1]
    Pairwise <- matrix(nrow = Num_Candidates, ncol = Num_Candidates)
    for (CurCand in 1:Num_Candidates) {
        CandRank <- as.vector(as.matrix(Votes[CurCand, ]))
        Pref_Cur_Cand <- t(Votes) - CandRank
        for (Pairs in 1:Num_Candidates) {
            Pairwise[CurCand, Pairs] <- sum(Pref_Cur_Cand[, Pairs] > 0)
        }
    }
    return(Pairwise)
}

Schulze <- function(PairsMatrix) {
    size <- dim(PairsMatrix)[1]
    p <- matrix(nrow = size, ncol = size)
    for (i in 1:size) {
        for (j in 1:size) {
            if (i != j) {
                if (PairsMatrix[i, j] > PairsMatrix[j, i]) {
                  p[i, j] <- PairsMatrix[i, j]
                } else {
                  p[i, j] <- 0
                }
            }
        }
    }
    for (i in 1:size) {
        for (j in 1:size) {
            if (i != j) {
                for (k in 1:size) {
                  if (i != k && j != k) {
                    p[j, k] <- max(p[j, k], min(p[j, i], p[i, k]))
                  }
                }
            }
        }
    }
    diag(p) <- 0
    return(p)
}

CondorcetRank <- function(BallotMatrix) {
    Num_Candidates <- dim(BallotMatrix)[1]
    Rankings <- matrix(nrow = Num_Candidates, ncol = 3)
    CurrentBallot <- BallotMatrix
    CurrentRank <- 1
    while (CurrentRank <= Num_Candidates) {
        CurrentNames <- as.vector(CurrentBallot[, 1])
        CurrentSize <- length(CurrentNames)
        CurrentVotes <- VoteExtract(CurrentBallot)
        Pairwise <- matrix(nrow = CurrentSize, ncol = CurrentSize)
        Pairwise <- PairCount(CurrentVotes)
        Winner <- vector(length = CurrentSize)

        # Check for Condorcet Winner

        for (i in 1:CurrentSize) {
            Winner[i] <- sum(Pairwise[i, ] > Pairwise[, i]) == (CurrentSize -
                1)
        }
        if (sum(Winner == TRUE) == 1) {
            # Condorcet Winner Exists
            CurrentWinner <- which(Winner == TRUE)
            Rankings[CurrentRank, ] <- c(CurrentNames[CurrentWinner], CurrentRank,
                "Condorcet")
        } else {

            # Condorcet Winner does not exist, calculate Schulze beatpaths

            Pairwise <- Schulze(Pairwise)
            for (i in 1:CurrentSize) {
                Winner[i] <- sum(Pairwise[i, ] > Pairwise[, i]) == (CurrentSize -
                  1)
            }
            if (sum(Winner == TRUE) == 1) {
                # Schwartz set has unique member
                CurrentWinner <- which(Winner == TRUE)
                Rankings[CurrentRank, ] <- c(CurrentNames[CurrentWinner], CurrentRank,
                  "Schulze")
            }
        }
        CurrentBallot <- CurrentBallot[-CurrentWinner, ]
        CurrentRank = CurrentRank + 1
    }
    Rankings <- data.frame(Rankings)
    names(Rankings) <- c("Name", "Rank", "Method")
    return(Rankings)
}
CondorcetRank(votes)
```

So here again, Candidate A is the winner, but now the runner-up is Candidate B:
that is, in all pairwise elections, Candidate B came out the winner more often
than Candidate D.

As it turned out, looking at the data, Candidate A had the most first
preferences anyway with three votes, so to keep it simple, we should have picked
them anyway!  But what's interesting, is that while Candidates D and E each had
2 first preferences, they were ranked third and fourth respectively by the
Condorcet method: most people preferred other candidates.

What if the preferences had been slightly different? Let's look at swapping two
preferences for a single judge: Judge 8 will now vote Candidate D first, and
Candidate E fourth:

```{r,echo=TRUE,eval=TRUE}
votes2 <- votes
votes2$J8[4] <- 1
votes2$J8[5] <- 4
```

The Borda method still places Candidate A first, with Candidates E and C
swapping places:

```{r,echo=FALSE,eval=TRUE}
AvgRank(votes2)
```

Most importantly though, the Concordet method now chooses Candidate D as the
winner, with Candidate A coming second!

```{r,echo=FALSE,eval=TRUE}
CondorcetRank(votes2)
```

So whilst Candidates A and D have the same number of first preferences, more
people preference Candidate D over all other candidates, than they do Candidate
A.

Majority rules!
